Prizm Silver Rookie Cards

### 2024-04-25 paper/paper.qmd
final touches

---
title: "Prizm Silver Rookie Cards"
subtitle: "An Analysis using statistic performance of a specific player and their rookie card price"
author: 
  - Harrison Huang
thanks: "Code and data are available at: https://github.com/lemonface88/NBA"
date: today 
date-format: long
abstract: "Collecting and investing in cards has been relevant for a while now. There are complex factors that contribute to the price fluxation of these cards. This analysis will use player statistics to create a model to try to grasp possible future price trends of the card. A linear model was used in this analysis. The results indicated that player statistics may still not be sufficient to predict future prices of the rookie card."
format: pdf
number-sections: true
bibliography: references.bib
---
```{r}
#| include: false
#| warning: false
#| message: false

@@

The analysis of these three datasets will be carried out using the statistical programming language `R` [@citeR], using the `tidyverse`[@citetidy], `here`[@citehere], and `readxl`[@citexl] packages. The figures and tables in the paper are generated using the, respectively, `ggplot2`[@citegg] and `knitr`[@citeknitr] packages.

@@

## Confounding variables
The linear model may be too simple due to each player's positions and playstyles. Because we are assuming that the higher the statistics are the better a player is performing. However, basketball is not only about statistics. If a player is known and in charge of defensive plays, their blocks and rebounds may be more important compared to attack-oriented players. The behavior of a player outside the league is also very crucial in terms of this. Even though Ja Morant played well in the league, he had multiple occasions where he displayed illegal objects during his livestream on Instagram. This devastated his reputation overall and he also received game restrictions as punishment. His card price also plummeted from around 2000 to now at around 400 dollars.

## Season timing
After taking a closer look at trends from both predicted and past data, an interesting point came up mind. When observing turning points of the price, they all seems to be the start and end of a season. Price increasing trends tends to happen at the start of each season, and then become more and more unpredictable throughout the season. This brings attention to the timing of the season and the price may be quite related and even more than the player statistics. Because despite implementing an increasing performance for each of the player statistics to run the model on, there are no clear signs of increasing price overall. The thing that stood out the most is the price increase at the start of every season in October. This could suggest that the initial hype of a new season could be contrbuting to the incresae of the card price.

@@

Future steps would include categorizing each player based on their positions and giving weight to the statistics that would have a better indication of their performance. For example, giving more weight to the average number of blocks per game for defensive players. Another step would be the incorporation of other models to find better fits, because of the complex relationship between the price and the player statistics, a more sophisticated model may be needed.

>>>>

---
title: "Prizm Silver Rookie Cards"
subtitle: "An Analysis Using Statistic Serformance of an Individual Player and Their Rookie Card Price"
author: 
  - Harrison Huang
thanks: "Code and data are available at: https://github.com/lemonface88/NBA"
date: today 
date-format: long
abstract: "Collecting and investing in cards has been relevant for a while now. There are complex factors that contribute to the price fluxation of these cards. This analysis will use player statistics to create a model to try to grasp possible future price trends of the card. A linear model was used in this analysis. The results indicated that player statistics may still not be sufficient to predict future prices of the rookie card."
format: pdf
number-sections: true
bibliography: references.bib
---
```{r}
#| include: false
#| warning: false
#| message: false

@@

The analysis of these three data sets will be carried out using the statistical programming language `R` [@citeR], using the `here`[@R-here], and `readxl`[@citexl] packages. Various packages were used throughout the writing of this paper including `tidyverse` [@tidyverse], `arrow` [@arrow], `here` [@here], `ggmap` [@ggmap], `knitr` [@knitr], `dplyr` [@dplyr], `janitor` [@janitor], `lintr` [@lintr], and `rstanarm` [@rstanarm].

@@

## Confounding variables
The linear model may be too simple due to each player's positions and play styles. Because we are assuming that the higher the statistics are the better a player is performing. However, basketball is not only about statistics. If a player is known and in charge of defensive plays, their blocks and rebounds may be more important compared to attack-oriented players. The mock data had in increase of 2% each month of each players statistics, which should lead to overall increase of the price if the statistics were all positively related to the price. The weight of an varibale might be overlooked or underlooked which could be casuing this issue.

The behavior of a player outside the league is also very crucial in terms of this. Even though Ja Morant played well in the league, he had multiple occasions where he displayed illegal objects during his live stream on Instagram. This devastated his reputation overall and he also received game restrictions as punishment. His card price also plummeted from around 2000 to now at around 400 dollars.

## Time and Price

The time frame also plays a crucial role in prices of many items in our every day life. For example, the rent price in Toronto starting May is lower compared to September from my past experience. This is due to the mass fluctuation of students leaving Toronto in May and coming back to Toronto in September. Timing and price in this case also plays a role in the change or card prices.

After taking a closer look at trends from both predicted and past data, an interesting point came up mind. When observing turning points of the price, they all seems to be the start and end of a season. Price increasing trends tends to happen at the start of each season, and then become more and more unpredictable throughout the season. This brings attention to the timing of the season and the price may be quite related and even more than the player statistics. Because despite implementing an increasing performance for each of the player statistics to run the model on, there are no clear signs of increasing price overall. The thing that stood out the most is the price increase at the start of every season in October. This could suggest that the initial hype of a new season could be contrbuting to the increase of the card price.

@@

Future steps would include categorizing each player based on their positions and giving weight to the statistics that would have a better indication of their performance. For example, giving more weight to the average number of blocks per game for defensive players. Upon taking a look at @fig-4, @fig-5, and @fig-6, the fit of the data are not very good. Another step would be the incorporation of other models to find better fits, because of the complex relationship between the price and the player statistics, a more sophisticated model may be needed. I think also using an average of all the player's statistics and price data and compare them to individual model would also be something beneficial, exploring the overall relationship rather than one at a time. Overall, this analysis gives insgitht to how complex the relationship is and that a lot more work would be required to be done.
